Copulas and their potential for ecology

### 2019-12-18 Paper.Rmd
further progress

The goal of this paper is to explore the potential for 
applications of copulas in ecology and to 
help determine to what extent ecological understanding may benefit generally from
using copulas. We provide a basic introduction
to copula concepts in the "Background on copulas" section below (section \ref{Background}). 
Readers can find additional abbreviated [@Anderson2018; @Genest2007] 
and comprehensive [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]
introductions elsewhere. Copulas have already been used very effectively 
in a few studies to illuminate ecological phenomena 
[@Valpine2014; @Anderson2018; @Popovic2018], but such usage is
still rare. Our results suggest that environmental, 
ecological and evolutionary processes 
may commonly generate and transmit complex dependence structures, including asymmetric tail 
associations, and that greater use of copulas can illuminate the underlying processes.
We believe copula approaches are among the tools all ecologists, including tropical
ecologists, should be considering for analysis of their data in the 21st century.

@@

<!--referee-requested para comparing to regression with heteroskedastic error term-->
Non-copula methods may be useful under some circumstances for exploring
asymmetries of tail association, but copula approaches have theoretical
and practical advantages. It was suggested to us that asymmetries of
tail association can also be explored using, for instance, a regression 
model with heteroskedastic error term, $y=mx+b+\epsilon(x)$. Whereas this 
approach seems capable of yielding insight for some datasets, it implicitly assumes
a particular causal relationship between the variable, an untenable assumption for our
synchrony applications and many other applications. Furthermore, no equivalent
of Sklar's theorem exists under this approach, so information on marginal 
distributions cannot be separated from association information. Sklar's theorem
means that the copula associated with two variables contains all the information
about the association. Such statements of 
mathematical completeness will not be avilable for others methods.

@@

We thank Joel E. Cohen for making us aware of copulas and their potential ecological utility in the 
first place. We thank the many contributors to the large datasets 
we used; D. Stevens and P. Verrier for data 
extraction; and David Tilman, Lauren Hallet, Jonathan Walter, Thomas 
Anderson, Lei Zhao, Andrew Rypel, and editors and annonymous reviewers for helpful suggestions. 
We thank James Bell of the Rothamsted Insect Survey (RIS). 
The RIS, a UK Capability, is funded by the Biotechnology and Biological Sciences Research Council 
under the Core Capability Grant BBS/E/C/000J0200. SG, LWS and DCR were partly funded by US National 
Science Foundation grants 1714195 and 1442595 and the James S McDonnell Foundation.

>>>>

The goal of this paper is to explore the potential for 
applications of copulas in ecology and to 
help determine to what extent ecological understanding may benefit generally from
using copulas. We provide a basic introduction
to copula concepts in the "Background on copulas" section below (section \ref{Background}). 
Readers can find additional abbreviated [@Anderson2018; @Genest2007] 
and comprehensive [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]
introductions elsewhere. Copulas have already been used very effectively 
in a few studies to illuminate ecological phenomena 
[@Valpine2014; @Anderson2018; @Popovic2018], but such usage is
still rare. Whereas @Anderson2018 provide a specific methodological pathway, based partly 
on copulas, for the important problem of the 
analysis of multivariate ecological population count data;
we instead advocate more broadly for the application of copulas across ecology.
Our results suggest that environmental, 
ecological and evolutionary processes 
may commonly generate and transmit complex dependence structures, including asymmetric tail 
associations, and that greater use of copulas can illuminate the underlying processes.
We believe copula approaches are among the tools all ecologists, including tropical
ecologists, should be considering for analysis of their data in the 21st century.

@@

<!--referee-requested para comparing to regression with heteroskedastic error term-->
Non-copula methods may be useful under some circumstances for exploring
asymmetries of tail association, but copula approaches have theoretical
and practical advantages. It was suggested to us that asymmetries of
tail association can also be explored using a regression 
model with heteroskedastic error term, $y=mx+b+\epsilon(x)$. Whereas this 
approach seems capable of yielding insight for some datasets, it implicitly assumes
a particular causal relationship between $x$ and $y$, an untenable assumption for 
many applications. Furthermore, no equivalent
of Sklar's theorem exists under this approach, so information on marginal 
distributions cannot be separated from association information, and the two types
of information will both influence regression results. Sklar's theorem
means that the copula associated with two variables contains all and only the information
about the association. Such statements of 
mathematical completeness will not be avilable for other methods.

@@

We thank Joel E. Cohen for making us aware of copulas in the 
first place. We thank the many contributors to the large datasets 
we used; D. Stevens and P. Verrier for data 
extraction; and David Tilman, Lauren Hallet, Jonathan Walter, Thomas 
Anderson, Lei Zhao, Andrew Rypel, and editors and annonymous reviewers for helpful suggestions. 
We thank James Bell of the Rothamsted Insect Survey (RIS). 
The RIS, a UK Capability, is funded by the Biotechnology and Biological Sciences Research Council 
under the Core Capability Grant BBS/E/C/000J0200. SG, LWS and DCR were partly funded by US National 
Science Foundation grants 1714195 and 1442595, the James S McDonnell Foundation, and the 
California Delta Science Program.
.
